---
title: "Analysis of CF_BAL_Pilot & C133_Neeland Data Combined"
subtitle: "Other Cells Sub-clustering"
author: "Jovana Maksimovic"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
site: workflowr::wflow_site
bibliography: ref.bib
output:
  bookdown::html_document2:
    base_format: workflowr::wflow_html
    toc: yes
    toc_float: yes
    theme: cosmo
    highlight: textmate
    number_sections: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


# Load libraries

```{r, message=FALSE, echo=FALSE}
suppressPackageStartupMessages(library(BiocStyle))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(glue))
suppressPackageStartupMessages(library(DropletUtils))
suppressPackageStartupMessages(library(scran))
suppressPackageStartupMessages(library(scater))
suppressPackageStartupMessages(library(scuttle))
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(patchwork))
suppressPackageStartupMessages(library(clustree))
suppressPackageStartupMessages(library(glmGamPoi))
suppressPackageStartupMessages(library(BiocParallel))
suppressPackageStartupMessages(library(paletteer))
suppressPackageStartupMessages(library(glmGamPoi))
suppressPackageStartupMessages(library(tidyHeatmap))
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(org.Hs.eg.db))
suppressPackageStartupMessages(library(EGSEA))
source(here("code/utility.R"))
source(here("code/helper_functions.R"))

set.seed(42)
options(scipen=-2)
options(future.globals.maxSize = 6500 * 1024^2)
```

# Load Data

Load the clustered and labelled CF_BAL_Pilot and C133_Neeland data.

```{r}
seu1 <- readRDS(file = here("data/SCEs/05_COMBO.clustered_annotated_lung_diet.SEU.rds"))
seu2 <- readRDS(file = here("data/SCEs/05_COMBO.clustered_annotated_others_diet.SEU.rds"))
seu <- merge(seu1, y = seu2)

DefaultAssay(seu) <- "RNA"
entrez <- select(org.Hs.eg.db, columns = c("ENTREZID","SYMBOL"), 
                 keys = keys(org.Hs.eg.db)) 
entrez <- entrez[!is.na(entrez$ENTREZID),]
seu <- seu[alias2SymbolTable(rownames(seu)) %in% entrez$SYMBOL,]
seu
```

# Subcluster other cells

Normalise and integrate data.

```{r}
out <- here("data/SCEs/06_COMBO.others_integrated.SEU.rds")

if(!file.exists(out)){ 
  seuInt <- intDat(seu, type = "RNA", 
                   reference = unique(seu$capture[seu$experiment == 1]),
                   #k.weight = min(table(seu$donor)))
                   k.weight = min(table(seu$donor)) - 1)
  saveRDS(seuInt, file = out)
  
} else {
  seuInt <- readRDS(file = out)
  
}
```

Visualise the data.

```{r}
seuInt <- RunPCA(seuInt, verbose = FALSE, dims = 1:30) %>%
  RunUMAP(verbose = FALSE, dims = 1:30)
```

```{r}
DimPlot(seuInt, group.by = "experiment", combine = FALSE)
```

# Clustering

## Perform Linear Dimensional Reduction

```{r}
p1 <- DimPlot(seuInt, reduction = "pca", group.by = "donor")
p2 <- DimPlot(seuInt, reduction = "pca", dims = c(1,3), group.by = "donor")
p3 <- DimPlot(seuInt, reduction = "pca", dims = c(2,3), group.by = "donor")
p4 <- DimPlot(seuInt, reduction = "pca", dims = c(3,4), group.by = "donor")

((p1 | p2) / (p3 | p4)) + plot_layout(guides = "collect") &
  theme(legend.text = element_text(size = 8),
        plot.title = element_text(size = 10),
        axis.title = element_text(size = 9),
        axis.text = element_text(size = 8))
```

```{r, fig.width=9, fig.height=27}
DimHeatmap(seuInt, dims = 1:30, cells = 500, balanced = TRUE)
```

## Determine the ‘Dimensionality’ of the Dataset

```{r}
ElbowPlot(seuInt, ndims = 30)
```

# Cluster the Cells

Examine cluster number and size with respect to resolution.

```{r, warning=FALSE, message=FALSE, fig.asp=1.3}
out <- here("data/SCEs/06_COMBO.others_clustered.SEU.rds")

if(!file.exists(out)){
  seuInt <- FindNeighbors(seuInt, reduction = "pca", dims = 1:30)
  seuInt <- FindClusters(seuInt, algorithm = 3, 
                         resolution = seq(0.1, 1, by = 0.1))
  seuInt <- RunUMAP(seuInt, dims = 1:10)
  saveRDS(seuInt, file = out)
  
} else {
  seuInt <- readRDS(file = out)
  
}

clustree::clustree(seuInt)
```

Choose a resolution. Visualise UMAP.

```{r, fig.asp=0.8}
grp <- "integrated_snn_res.1"
DimPlot(seuInt, reduction = 'umap', label = TRUE, repel = TRUE, 
        label.size = 2.5, group.by = grp) + NoLegend()
```

```{r, fig.asp=1}
options(ggrepel.max.overlaps = Inf)
DimPlot(seuInt, reduction = 'umap', label = TRUE, repel = TRUE, 
        label.size = 2.5, group.by = "predicted.ann_level_3") +
  theme(legend.position = "bottom") + NoLegend()
```


## Examine clusters

Visualise quality metrics by cluster.

```{r, fig.asp=0.8}
seuInt@meta.data %>%
  ggplot(aes(x = integrated_snn_res.1,
             y = predicted.annotation.l1.score,
             fill = integrated_snn_res.1)) +
  geom_violin(scale = "width") +
  NoLegend() -> p1

seuInt@meta.data %>%
  ggplot(aes(x = integrated_snn_res.1,
             y = nCount_RNA,
             fill = integrated_snn_res.1)) +
  geom_violin(scale = "area") +
  scale_y_log10() +
  NoLegend() -> p2

seuInt@meta.data %>%
  ggplot(aes(x = integrated_snn_res.1,
             y = nFeature_RNA,
             fill = integrated_snn_res.1)) +
  geom_violin(scale = "area") +
  scale_y_log10() +
  NoLegend() -> p3

seuInt@meta.data %>%
  ggplot(aes(x = integrated_snn_res.1,
             y = predicted.ann_level_3.score,
             fill = integrated_snn_res.1)) +
  geom_violin(scale = "area") +
  scale_y_log10() +
  NoLegend() -> p4

((p1 | p2) / (p3 | p4)) & theme(text = element_text(size = 8))
```

# Identify Cluster Marker Genes

Adapted from Dr. Belinda Phipson's [work](https://bphipson.github.io/Human_Development_snRNAseq/14-MarkerAnalysisBCT.html) for [@Sim2021-cg].

## Test for Marker Genes using *limma*

```{r}
# limma-trend for DE
Idents(seuInt) <- grp
counts <- as.matrix(seuInt[["RNA"]]@counts)

y.org <- DGEList(counts)
logcounts <- normCounts(y.org, log = TRUE, prior.count = 0.5)

maxclust <- length(levels(Idents(seuInt))) - 1

clustgrp <- paste0("c", Idents(seuInt))
clustgrp <- factor(clustgrp, levels = paste0("c", 0:maxclust))
donor <- seuInt$donor

design <- model.matrix(~ 0 + clustgrp + donor)
colnames(design)[1:(length(levels(clustgrp)))] <- levels(clustgrp)

# Create contrast matrix
mycont <- matrix(NA, ncol = length(levels(clustgrp)), 
                 nrow = length(levels(clustgrp)))
rownames(mycont) <- colnames(mycont) <- levels(clustgrp)
diag(mycont) <- 1
mycont[upper.tri(mycont)] <- -1/(length(levels(factor(clustgrp))) - 1)
mycont[lower.tri(mycont)] <- -1/(length(levels(factor(clustgrp))) - 1)

# Fill out remaining rows with 0s
zero.rows <- matrix(0, ncol = length(levels(clustgrp)),
                    nrow = (ncol(design) - length(levels(clustgrp))))
fullcont <- rbind(mycont, zero.rows)
rownames(fullcont) <- colnames(design)

fit <- lmFit(logcounts, design)
fit.cont <- contrasts.fit(fit, contrasts = fullcont)
fit.cont <- eBayes(fit.cont, trend = TRUE, robust = TRUE)

summary(decideTests(fit.cont))
```

## Test relative to a threshold (TREAT)

```{r}
tr <- treat(fit.cont, fc = 1.5)
dt <- decideTests(tr)
summary(dt)
```

### Mean-difference Plots per Cluster

```{r, fig.asp=1}
par(mfrow=c(3,3))

for(i in 1:ncol(mycont)){
  plotMD(tr, coef = i, status = dt[,i], hl.cex = 0.5)
  abline(h = 0, col = "lightgrey")
  lines(lowess(tr$Amean, tr$coefficients[,i]), lwd = 1.5, col = 4)
}
```


                
### Export Marker Genes per cluster

```{r}
options(scipen=-1, digits = 6)
contnames <- colnames(mycont)
dirName <- here("output/marker-analysis/05-COMBO-others")
if(!dir.exists(dirName)) dir.create(dirName)

getCols <- setNames(c("SYMBOL","ENTREZID"),c("SYMBOL","ENTREZID"))
tr$genes <- data.frame(
  lapply(getCols, function(column) {
    mapIds(
      x = org.Hs.eg.db,
      keys = rownames(tr),
      keytype = "SYMBOL",
      column = column)
  }),
  row.names = rownames(tr))

gsAnnots <- buildIdx(entrezIDs = tr$genes$ENTREZID, species = "human",
                     msigdb.gsets = c("c2","c5"))
reactomeIdx <-gsAnnots$c2@idx[grep("REACTOME", 
                                   names(gsAnnots$c2@idx))]

for(i in 1:length(contnames)){
  top <- topTreat(tr, coef = i, n = Inf)
  top <- top[top$logFC > 0, ]
  
  write.csv(top[1:100, ],
            file = glue("{dirName}/up-cluster-{contnames[i]}.csv"))
  
  cameraPR(tr$t[,i], reactomeIdx) %>%
    rownames_to_column(var = "Pathway") %>%
    slice_head(n = 20) %>%
    write_csv(file = here(glue("{dirName}/REACTOME-cluster-{contnames[i]}.csv")))
}
```

### Cluster marker gene dot plot

Genes duplicated between clusters are excluded.

```{r, message=FALSE, warning=FALSE, fig.asp=3}
sig.genes <- vector("list", ncol(tr))
p <- vector("list",length(sig.genes))
DefaultAssay(seuInt) <- "RNA"

for(i in 1:length(sig.genes)){
  top <- topTreat(tr, coef = i, n = Inf)
  sig.genes[[i]] <- rownames(top)[top$logFC > 0][1:10]
}

sig <- unlist(sig.genes)
geneCols <- c(rep(rep(c("grey","black"), each = 10), ncol(tr)/2), 
              rep("grey", 10))[!duplicated(sig)] 

DotPlot(seuInt,    features = sig[!duplicated(sig)], 
                    group.by = "integrated_snn_res.1",
                    cols = c("lightgrey", "red"),
                    dot.scale = 3) + 
    RotatedAxis() + 
    FontSize(y.text = 8, x.text = 12) + 
    labs(y = element_blank(), x = element_blank()) + 
    coord_flip() + 
  theme(axis.text.y = element_text(color = geneCols)) +
  ggtitle("Top 10 cluster marker genes without duplicates")
```

### Neutrophil relevant gene expression

```{r, fig.width=11, fig.asp=0.4}
neuMarkers <- c("CSF3R","FPR1","FCGR3B","NAMPT","MNDA","S100A8","FUT4","CEACAM8",
                "PLAUR","APOBEC3A","SRGN","AIF1","IL1RN","IF1B","SOD2","FCN1")

DoHeatmap(seuInt,
          group.by = "integrated_snn_res.1", size = 2.5,
          features = neuMarkers, assay = "RNA", slot = "data") + 
  NoLegend() +
  theme(axis.text = element_text(size = 6))
```

### No. cells per cluster

```{r}
seuInt@meta.data %>%
  ggplot(aes(x = integrated_snn_res.1, fill = integrated_snn_res.1)) +
  geom_bar() +
  geom_text(aes(label = ..count..), stat = "count", 
            vjust = -0.5, colour = "black", size = 2) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  NoLegend()
```
# Load protein data
## Add to `Seurat` object

```{r}
seuAdt <- readRDS(here("data",
                       "SCEs",
                       "05_COMBO.clustered_annotated_adt_diet.SEU.rds"))
seuAdt <- subset(seuAdt, cells = colnames(seuInt))
all(colnames(seuAdt) == colnames(seuInt))

seuInt[["ADT.dsb"]] <- seuAdt[["ADT.dsb"]]
seuInt[["ADT.raw"]] <- seuAdt[["ADT.raw"]]
seuInt
```

```{r}
rm(seuAdt)
gc()
```

## Load protein annotations

```{r, message=FALSE, warning=FALSE}
prots <- read.csv(file = here("data",
                              "sample_sheets",
                              "TotalSeq-A_Universal_Cocktail_v1.0.csv")) %>%
  dplyr::filter(grepl("^A0", id)) %>%
  dplyr::filter(!grepl("[Ii]sotype", name)) 
```

## Visualise all ADTs

Normalised with DSB. CITE-seq ADT data was transferred to scRNA-seq using reference mapping and transfer.

```{r, fig.asp=1}
cbind(seuInt@meta.data, 
      as.data.frame(t(seuInt@assays$ADT.dsb@data))) %>% 
  dplyr::group_by(integrated_snn_res.1, experiment) %>% 
  dplyr::summarize_at(.vars = prots$id, .funs = median) %>%
  pivot_longer(c(-integrated_snn_res.1, -experiment), names_to = "ADT",
               values_to = "ADT Exp.") %>%
  left_join(prots, by = c("ADT" = "id")) %>%
  mutate(Cluster = as.character(integrated_snn_res.1)) %>%
  dplyr::rename(Protein = name) |> 
  dplyr::filter(experiment == 2) |>
  ungroup() -> dat

plot(density(dat$`ADT Exp.`))
topMax <- 8
abline(v = topMax, lty = 2, col = "grey")
```

```{r, fig.asp=2.5}
dat |>
  heatmap(
    .column = Cluster,
    .row = Protein,
    .value = `ADT Exp.`,
    scale = "none",
    palette_value = circlize::colorRamp2(seq(-1, topMax, length.out = 256),
                                         viridis::magma(256)),
    rect_gp = grid::gpar(col = "white", lwd = 1),
    show_row_names = TRUE,
    column_names_gp = grid::gpar(fontsize = 10),
    column_title_gp = grid::gpar(fontsize = 12),
    row_names_gp = grid::gpar(fontsize = 8),
    row_title_gp = grid::gpar(fontsize = 12),
    column_title_side = "top",
    heatmap_legend_param = list(direction = "vertical")) 
```

## Visualise ADTs of interest

```{r}
adt <- read_csv(file = here("data/Proteins_other_22.04.22.csv"))
adt <- adt[!duplicated(adt$DNA_ID),]

dat %>%
  inner_join(adt, by = c("ADT" = "DNA_ID")) %>%
  dplyr::mutate(Protein = `Name for heatmap`) |> 
  heatmap(
    .column = Cluster,
    .row = Protein,
    .value = `ADT Exp.`,
    scale = "none",
    palette_value = circlize::colorRamp2(seq(-1, topMax, length.out = 256),
                                         viridis::magma(256)),
    rect_gp = grid::gpar(col = "white", lwd = 1),
    show_row_names = TRUE,
    column_names_gp = grid::gpar(fontsize = 10),
    column_title_gp = grid::gpar(fontsize = 12),
    row_names_gp = grid::gpar(fontsize = 8),
    row_title_gp = grid::gpar(fontsize = 12),
    column_title_side = "top",
    heatmap_legend_param = list(direction = "vertical")) 
```

## Visualise cytokines of interest

```{r}
markers <- read_csv(file = here("data", 
                                "other_subclusters_cytokines.csv"),
                    col_names = FALSE)
p <- DotPlot(seuInt, 
             features = markers$X1,
             cols = c("grey", "red"),
             dot.scale = 5,
             assay = "RNA",
             group.by = "integrated_snn_res.1") +
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1, 
                                   vjust = 0.5,
                                   size = 8),
        axis.text.y = element_text(size = 8),
        text = element_text(size = 8)) +
  coord_flip() +
  labs(y = "Cluster", x = "Cytokine")

p
```


# Session info {.appendix}

<summary>The analysis and this document were prepared using the following software (click triangle to expand)</summary>
<details>

```{r}
sessioninfo::session_info()
```

# References
