---
title: "Analysis of scRNA-seq & CITE-seq Data Combined"
subtitle: "Post-process T/NK Cells Sub-clustering Results"
author: "Jovana Maksimovic"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
site: workflowr::wflow_site
bibliography: ref.bib
output:
  bookdown::html_document2:
    base_format: workflowr::wflow_html
    toc: yes
    toc_float: yes
    theme: cosmo
    highlight: textmate
    number_sections: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Load libraries

```{r, message=FALSE, echo=FALSE}
suppressPackageStartupMessages(library(BiocStyle))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(glue))
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(patchwork))
suppressPackageStartupMessages(library(paletteer))
suppressPackageStartupMessages(library(tidyHeatmap))
suppressPackageStartupMessages(library(speckle))
suppressPackageStartupMessages(library(Cepo))
source(here("code/utility.R"))
source(here("code/helper_functions.R"))

set.seed(42)
options(scipen=999)
options(future.globals.maxSize = 6500 * 1024^2)
```

# Load Data

```{r, warning=FALSE, message=FALSE, fig.asp=1.25}
out <- here("data/SCEs/06_COMBO.tcells_clustered.SEU.rds")
seuInt <- readRDS(file = out)

seuInt
```
# Sub-cluster annotation
## Load manual annotations

```{r}
labels <- read_csv(here("data/TNK_subcluster_annotation_29.05.22.csv"))

seuInt@meta.data %>%
  left_join(labels %>%
              mutate(Annotation = ifelse(is.na(Annotation),
                                         "SUSPECT",
                                         Annotation),
                     Broad = ifelse(is.na(Broad),
                                         "SUSPECT",
                                         Broad)) %>%
              mutate(Cluster = as.factor(Cluster),
                     Annotation = as.factor(Annotation)),
            by = c("integrated_snn_res.1" = "Cluster")) -> ann 

ann %>% dplyr::pull(Annotation) -> seuInt$Annotation
ann %>% dplyr::pull(Broad) -> seuInt$Broad

seuInt$Annotation <- fct_drop(seuInt$Annotation)
seuInt$Broad <- fct_drop(seuInt$Broad)
seuInt
```

## Visulalise annotations

```{r, fig.width=10, fig.asp=0.4}
options(ggrepel.max.overlaps = Inf)
DimPlot(seuInt, reduction = 'umap', label = TRUE, repel = TRUE, 
        label.size = 2.5, group.by = "integrated_snn_res.1") + 
  NoLegend() -> p1
DimPlot(seuInt, reduction = 'umap', label = TRUE, repel = TRUE, 
        label.size = 2.5, group.by = "Annotation") + 
  NoLegend() +
  scale_color_paletteer_d("miscpalettes::pastel") -> p2
DimPlot(seuInt, reduction = 'umap', label = TRUE, repel = TRUE, 
        label.size = 2.5, group.by = "Broad") + 
  NoLegend() +
  scale_color_paletteer_d("miscpalettes::pastel") -> p3

(p1 | p2 | p3) & theme(text = element_text(size = 8),
                  axis.text = element_text(size = 8))
```

### No. cells per cluster

```{r}
seuInt@meta.data %>%
  ggplot(aes(x = Annotation, fill = Annotation)) +
  geom_bar() +
  geom_text(aes(label = ..count..), stat = "count",
            vjust = -0.5, colour = "black", size = 2) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  NoLegend() +
  scale_fill_paletteer_d("miscpalettes::pastel")

seuInt@meta.data %>%
  ggplot(aes(x = Broad, fill = Broad)) +
  geom_bar() +
  geom_text(aes(label = ..count..), stat = "count",
            vjust = -0.5, colour = "black", size = 2) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  NoLegend() +
  scale_fill_paletteer_d("miscpalettes::pastel")
```

### Remove SUSPECT clusters

```{r}
seuInt <- subset(seuInt, cells = which(seuInt$Annotation != "SUSPECT")) 
seuInt$Annotation <- fct_drop(seuInt$Annotation)
seuInt$Broad <- fct_drop(seuInt$Broad)

DefaultAssay(seuInt) <- "integrated"
seuInt <- RunPCA(seuInt, verbose = FALSE, dims = 1:30) %>%
  RunUMAP(verbose = FALSE, dims = 1:30)
```

## Visualise annotations

```{r}
DimPlot(seuInt, reduction = 'umap', label = TRUE, repel = FALSE, 
        label.size = 3, group.by = "Annotation") + 
  NoLegend() +
  scale_color_paletteer_d("miscpalettes::pastel") +
  xlim(c(-6, 8)) -> p2

DimPlot(seuInt, reduction = 'umap', label = TRUE, repel = FALSE, 
        label.size = 3, group.by = "Broad") + 
  NoLegend() +
  scale_color_paletteer_d("miscpalettes::pastel") -> p1

(p1) & theme(text = element_text(size = 8),
                  axis.text = element_text(size = 8))

(p2) & theme(text = element_text(size = 8),
                  axis.text = element_text(size = 8)) -> f3a
f3a
```

## `Cepo` cluster marker genes 

```{r}
cepoMarkers <- Cepo(seuInt[["RNA"]]@data, 
                   seuInt$Annotation, 
                   exprsPct = 0.1,
                   logfc = 1)

sapply(1:ncol(cepoMarkers$stats), function(i){
  names(sort(cepoMarkers$stats[,i], decreasing = TRUE))[1:20]
}) -> dat

colnames(dat) <- colnames(cepoMarkers$stats)
dat %>% knitr::kable()
```

### `Cepo` marker gene dot plot

Genes duplicated between clusters are excluded.

```{r, fig.width=9, fig.asp=0.4}
DefaultAssay(seuInt) <- "RNA"

maxGenes <- 5
sigGenes <- lapply(1:ncol(dat), function(i){
  dat[,i][1:maxGenes]
})
sig <- unlist(sigGenes)
geneCols <- c(rep(rep(c("blue","black"), each = maxGenes), 
                  ceiling(ncol(dat)/2)))[1:length(sig)][!duplicated(sig)] 

geneCols <- rep(paletteer_d("miscpalettes::pastel", ncol(dat)), 
                each = maxGenes)[1:length(sig)][!duplicated(sig)] 

pal <- paletteer::paletteer_d("vapoRwave::cool")
DotPlot(seuInt,    
        features = sig[!duplicated(sig)], 
        group.by = "Annotation",
        dot.scale = 2.5) + 
  FontSize(y.text = 10, x.text = 9) + 
  labs(y = element_blank(), x = element_blank()) + 
  theme(axis.text.x = element_text(color = geneCols,
                                   angle = 90,
                                   hjust = 1,
                                   vjust = 0.5,
                                   face = "bold"),
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 10)) +
  scale_color_gradient2(low = pal[1], 
                        mid = pal[3], 
                        high = pal[5]) -> f3c
f3c
```

## Visualise cytokines of interest

```{r, fig.asp=0.8}
markers <- read_csv(file = here("data",
                                "T-NK_subclusters_cytokines.csv"),
                    col_names = FALSE)
p <- DotPlot(seuInt,
             features = markers$X1,
             cols = c("grey", "red"),
             dot.scale = 5,
             assay = "RNA",
             group.by = "Annotation") +
  theme(axis.text.x = element_text(angle = 90,
                                   hjust = 1,
                                   vjust = 0.5,
                                   size = 8),
        axis.text.y = element_text(size = 8),
        text = element_text(size = 8)) +
  coord_flip() +
  labs(y = "Label", x = "Cytokine")

p
```

# Load protein data
## Add to `Seurat` object

```{r}
seuAdt <- readRDS(here("data",
                       "SCEs",
                       "04_COMBO.clustered_annotated_adt_diet.SEU.rds"))
seuAdt <- subset(seuAdt, cells = colnames(seuInt))
all(colnames(seuAdt) == colnames(seuInt))

seuInt[["ADT.dsb"]] <- seuAdt[["ADT.dsb"]]
seuInt[["ADT.raw"]] <- seuAdt[["ADT.raw"]]
seuInt
```
```{r}
rm(seuAdt)
gc()
```

## Load protein annotations

```{r, message=FALSE, warning=FALSE}
prots <- read_csv(file = here("data",
                              "sample_sheets",
                              "TotalSeq-A_Universal_Cocktail_v1.0.csv")) %>%
  dplyr::filter(grepl("^A0", id)) %>%
  dplyr::filter(!grepl("[Ii]sotype", name)) 
```
## Visualise all ADTs

Normalised with DSB. C133_Neeland ADT data was transferred to CF_BAL_Pilot using reference mapping and transfer.

```{r, fig.asp=1}
cbind(seuInt@meta.data, 
      as.data.frame(t(seuInt@assays$ADT.dsb@data))) %>% 
  dplyr::group_by(Annotation, experiment) %>% 
  dplyr::summarize_at(.vars = prots$id, .funs = median) %>%
  pivot_longer(c(-Annotation, -experiment), names_to = "ADT",
               values_to = "ADT Exp.")  %>%
  left_join(prots, by = c("ADT" = "id")) %>%
  mutate(`Cell Label` = Annotation) %>%
  dplyr::rename(Protein = name) |> 
  dplyr::filter(experiment == 2) |>
  ungroup() -> dat

plot(density(dat$`ADT Exp.`))
topMax <- 8
abline(v = topMax, lty = 2, col = "grey")
```

```{r, fig.asp=2.5}
dat |> heatmap(
  .column = `Cell Label`,
  .row = Protein,
  .value = `ADT Exp.`,
  scale = "none",
  rect_gp = grid::gpar(col = "white", lwd = 1),
  show_row_names = TRUE,
  column_names_gp = grid::gpar(fontsize = 10),
  column_title_gp = grid::gpar(fontsize = 12),
  row_names_gp = grid::gpar(fontsize = 8),
  row_title_gp = grid::gpar(fontsize = 12),
  column_title_side = "top",
    palette_value = circlize::colorRamp2(seq(-1, topMax, length.out = 256),
                                         viridis::magma(256)),
  heatmap_legend_param = list(direction = "vertical")) 
```

## Visualise ADTs of interest

```{r}
adt <- read_csv(file = here("data/Proteins_T-NK_22.04.22.csv"))
adt <- adt[!duplicated(adt$DNA_ID),]
  
dat |>
  dplyr::inner_join(adt, by = c("ADT" = "DNA_ID")) |>
  mutate(Protein = `Name for heatmap`) |>
  heatmap(
    .column = Protein,
    .row = `Cell Label`,
    .value = `ADT Exp.`,
    scale = "none",
    rect_gp = grid::gpar(col = "white", lwd = 1),
    show_row_names = TRUE,
    column_names_gp = grid::gpar(fontsize = 10),
    column_title_gp = grid::gpar(fontsize = 12),
    row_names_gp = grid::gpar(fontsize = 10),
    row_title_gp = grid::gpar(fontsize = 12),
    column_title_side = "bottom",
    heatmap_legend_param = list(direction = "vertical"),
    palette_value = circlize::colorRamp2(seq(-1, topMax, length.out = 256),
                                         viridis::magma(256)),
    column_title_side = "bottom") |>
  add_tile(`Cell Label`, show_legend = FALSE,
           show_annotation_name = FALSE,
           palette = paletteer_d("miscpalettes::pastel", 
                                 length(levels(seuInt$Annotation)))) -> f3d

wrap_heatmap(f3d)
```

# Proportions analysis
## Load clinical information

Import clinical characteristics and patient information and associate with `genetic_donor` IDs.

```{r}
info <- read.csv(file = here("data/sample_sheets/Sample_information.csv"))
tab <- table(seuInt$HTO, seuInt$donor)
tab <- tab[rownames(tab) != "Unknown",]

apply(tab, 2, function(x){
  names(which(x == max(x)))
}) %>% data.frame %>%
  dplyr::rename("HTO" = ".") %>%
  rownames_to_column(var = "donor") %>%
  inner_join(info, by = c("HTO" = "Sample")) %>%
  mutate(Batch = factor(Batch)) -> info

info %>% knitr::kable()
```

## Sub-cluster proportions (Broad)

```{r}
# Differences in cell type proportions
props <- getTransformedProps(clusters = seuInt$Broad,
                             sample = seuInt$donor, transform="asin")
props$Proportions %>% knitr::kable()
```

### Cell proportions by donor

```{r, fig.asp=1}
props$Proportions %>%
  data.frame %>%
  inner_join(info, by = c("sample" = "donor")) %>%
ggplot(aes(x = Participant, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust = 1),
        legend.text = element_text(size = 8)) +
  labs( y = "Proportion", fill = "Cell label") +
  scale_fill_paletteer_d("miscpalettes::pastel")
``` 


## Sub-cluster proportions (Fine)

```{r}
# Differences in cell type proportions
props <- getTransformedProps(clusters = seuInt$Annotation,
                             sample = seuInt$donor, transform="asin")
props$Proportions %>% knitr::kable()
```

### Cell proportions by donor

```{r, fig.asp=1}
props$Proportions %>%
  data.frame %>%
  inner_join(info, by = c("sample" = "donor")) %>%
ggplot(aes(x = Participant, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust = 1),
        legend.text = element_text(size = 8)) +
  labs(y = "Proportion", fill = "Cell label") +
  scale_fill_paletteer_d("miscpalettes::pastel") -> f3b
f3b
``` 


# Save data

```{r}
out <- here(glue("data/SCEs/06_COMBO.clean_tcells_diet.SEU.rds"))
if(!file.exists(out)){
  DefaultAssay(seuInt) <- "RNA"
  saveRDS(DietSeurat(seuInt,
                     assays = c("RNA", "ADT.dsb", "ADT.raw"),
                     dimreducs = NULL,
                     graphs = NULL), out)

}
```

# Panel figures

```{r, fig.width=12, fig.asp=1.3}
layout = "AAAB
          AAAB
          AAAB
          CCCC
          CCCC
          DDDD
          DDDD
          DDDD"
((f3a + ggtitle("")) + 
   f3b + 
   f3c +
   wrap_heatmap(f3d)) + 
  plot_layout(design = layout) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 14, face = "bold"))
```

# Session info {.appendix}

<summary>The analysis and this document were prepared using the following software (click triangle to expand)</summary>
<details>

```{r}
sessioninfo::session_info()
```

</details>
