---
title: "Getting Started"
description: |
author:
  - name: Jovana Maksimovic
date: "`r Sys.Date()`"
bibliography: ref.bib
site: workflowr::wflow_site
output:
  html_document:
    theme: cosmo
---

```{r knitr, include = FALSE}
knitr::opts_chunk$set(autodep        = TRUE,
                      cache          = FALSE,
                      cache.comments = FALSE,
                      echo           = TRUE,
                      error          = FALSE,
                      fig.align      = "center",
                      fig.width      = 10,
                      fig.height     = 8,
                      message        = FALSE,
                      warning        = FALSE)
```

This page describes how to download the data and code used in this analysis,
set up the project directory and reproduce the analysis. We have used the
[`workflowr`][workflowr] and [`renv`][renv] packages to organise this project and 
ensure reproducibility.

# Getting the code {#code}

All the code and outputs of this analysis are available from GitHub at
https://github.com/Oshlack/paed-cf-cite-seq. If you want to replicate the
analysis you can either clone the repository or download it as a zipped 
directory.

Once you have a local copy of the repository you should see the following
directory structure:

* `analysis/` - Contains the RMarkdown documents with the various stages of
analysis. These are numbered according to the order they should be run.
* `code/` - R scripts with custom functions used in some analysis stages. 
* `data/` - This directory contains the data files used in the analysis with
sub-directories for different data types (see [Getting the data](#data) for
details). Processed intermediate data files will also be placed here.
* `docs/` - This directory contains the analysis website html files hosted at
  http://oshlacklab.com/paed-cf-cite-seq, as well as the image files.
* `output/` - Directory for output files produced by the analysis.
* `renv/` 
* `README.md` - README describing the project.
* `.Rprofile` - Custom R profile for the project including set up for
  `workflowr`.
* `.gitattributes`
* `.gitmodules`
* `.gitignore` - Details of files and directories that are excluded from the
  repository.
* `.renvignore` - `renv` ignore file
* `_workflowr.yml` - `workflowr` configuration file.
* `paed-cf-cite-seq.Rproj` - RStudio project file.
* `renv.lock` - `renv` lock file, used to restore and install correct versions of
R packages required for this project.

This analysis was completed using `r R.version.string`. To ensure reproducibility
the `renv` package was used to track package sources and versions. Ensure you 
have the correct version of R and `renv` installed prior to beginning. To 
install the necessary package versions you can use:

```r
renv::restore()
```

For more information on using `renv` see the `renv` [website][renv].


# Getting the data {#data}

The raw counts generated for this study can be downloaded as RDS files from [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.4706148.svg)](https://doi.org/10.5281/zenodo.4706148).

To use the RDS objects, after cloning or downloading the GitHub repository to 
your computer, please extract the `raw_counts.tar.gz` archive under the 
`data/SCEs` directory, using: 

```r
tar -xvf raw_counts.tar.gz.
```

In this project we have also used publicly available single cell RNA-seq data 
generated from RBC-depleted cells from non-small cell lung tumor and the blood 
of 7 patients. The raw count data and metadata can be downloaded from 
[GSE127465][geo]. The `GSE127465_RAW.tar` and 
`GSE127465_human_cell_metadata_54773x25.tsv.gz` are required. The downloaded 
`tar` file should be extracted under the `data` directory by running the 
following command:

```r
tar –xvf GSE127465_RAW.tar
```

The `GSE127465_human_cell_metadata_54773x25.tsv.gz` should be placed in the 
newly created `GSE127465_RAW` directory.

The downstream analysis code assumes the following directory structure inside 
the `data/` directory:

* `GSE127465_RAW`
    * `GSE127465_human_cell_metadata_54773x25.tsv.gz`
    * `GSM3635278_human_p1t1_raw_counts.tsv.gz`
    * ...
    * ...
    * `GSM3635303_human_p7b1_raw_counts.tsv.gz`


# Running the analysis {#analysis}

The analysis directory contains the following analysis files:

```{r, echo=FALSE}
list.files(path = here::here("analysis"), pattern = "[01][0-9]_*")
```

As indicated by the numbering they should be run in this order. If you want to
reproduce the entire analysis this can be easily done using `workflowr`.

```r
workflowr::wflow_build(republish = TRUE)
```

It is also possible to run individual stages of the analysis, either by
providing the names of the file you want to run to `workflowr::wflow_build()` or
by manually knitting the document (for example using the 'Knit' button in
RStudio). Note, most parts of the analysis require outputs generated by a 
previous step and so will not run unless the preceding steps have already been 
executed.

[geo]: <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE127465> "geo"
[workflowr]: <https://workflowr.github.io/workflowr/> "workflowr"
[renv]: <https://rstudio.github.io/renv/articles/renv.html> "renv"
