---
title: "Getting Started"
description: |
author:
  - name: Jovana Maksimovic
date: "`r Sys.Date()`"
bibliography: ref.bib
site: workflowr::wflow_site
output:
  html_document:
    theme: cosmo
---

```{r knitr, include = FALSE}
knitr::opts_chunk$set(autodep        = TRUE,
                      cache          = FALSE,
                      cache.comments = FALSE,
                      echo           = TRUE,
                      error          = FALSE,
                      fig.align      = "center",
                      fig.width      = 10,
                      fig.height     = 8,
                      message        = FALSE,
                      warning        = FALSE)
```

This page describes how to download the data and code used in this analysis,
set up the project directory and rerun the analysis. We have used the
[`workflowr`][workflowr] package to organise the analysis and incorporate
reproducibilty information into the output documents.

# Getting the code {#code}

All the code and outputs of analysis are available from GitHub at
https://github.com/Oshlack/methyl-geneset-testing. If you want to replicate the
analysis you can either fork the repository and clone it or download the
repository as a zipped directory.

Once you have a local copy of the repository you should see the following
directory structure:

* `analysis/` - Contains the RMarkdown documents with the various stages of
analysis. These are numbered according to the order they should be run.
* `data/` - This directory contains the data files used in the analysis with
sub-directories for different data types (see [Getting the data](#data) for
details). Processed intermediate data files will also be placed here.
* `output/` - Directory for output files produced by the analysis, each analysis
step has it's own sub-directory.
* `docs/` - This directory contains the analysis website hosted at
  http://oshlacklab.com/methyl-geneset-testing, including image files.
* `code/` - R scripts with custom functions used in some analysis stages. There
  are sub-directories for scripts associated with different steps in the
  anlaysis.
* `README.md` - README describing the project.
* `.Rprofile` - Custom R profile for the project including set up for
  `workflowr`.
* `.gitignore` - Details of files and directories that are excluded from the
  repository.
* `_workflowr.yml` - Workflowr configuration file.
* `paed-cf-cite-seq.Rproj` - RStudio project file.


# Getting the data {#data}


# Running the analysis {#analysis}

The analysis directory contains the following analysis files:

```{r, echo=FALSE}
list.files(path = here::here("analysis"), pattern = "[01][0-9]_*")
```

As indicated by the numbering they should be run in this order. If you want to
rerun the entire analysis this can be easily done using `workflowr`.

```r
workflowr::wflow_build(republish = TRUE)
```

<!-- It is important to consider the computer and environment you are using before -->
<!-- doing this. Running this analysis from scratch requires a considerable amount -->
<!-- of time, disk space and memory. **Some stages of the analysis need to be executed -->
<!-- on a HPC to generate results required by downstream steps.** If you do no have -->
<!-- access to a HPC to perform these analyses using the code provided, you can -->
<!-- download pre-computed RDS files containing the results from [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.4706148.svg)](https://doi.org/10.5281/zenodo.4706148). -->

<!-- To use the pre-computed RDS objects, after cloning or downloading the GitHub repository to your computer, please extract the `outputs.tar.gz` archive under the `output` directory, using `tar -xvf outputs.tar.gz`. -->

It is also possible to run individual stages of the analysis, either by
providing the names of the file you want to run to `workflowr::wflow_build()` or
by manually knitting the document (for example using the 'Kint' button in
RStudio).

[expt-hub]: https://bioconductor.org/packages/release/bioc/html/ExperimentHub.html "ExperimentHub"
[geo-rnaseq]: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE107011 "GSE107011"
[sra-rnaseq]: https://www.ncbi.nlm.nih.gov/sra?term=SRP125125 "SRP125125"
[workflowr]: https://jdblischak.github.io/workflowr/ "Workflowr"


```{r session-info, cache = FALSE, eval = FALSE}
devtools::session_info()
```
